---
title: "PS5-md-plots"
author: "Kaitlyn Billington"
date: "2023-02-18"
output: html_document
---
# Load and check data (5pt)
Your first task is to do a very simple data check: 
1. (1pt) Create an rmarkdown file
_Note:_ This is that file
```{r, message=FALSE}
library(tidyverse)
```

2. (2pt) Load data. How many rows/columns do we have?
```{r}
gapminder <- read.delim("data/gapminder.csv.bz2")
dim(gapminder)
```


3. (2pt) Print a small sample of data. Does it look OK?
```{r}
gapminder %>% 
  sample_n(5)
```
There are many blanks/NAs, but this makes sense with the information being considered. The data frame itself looks good and is properly formatted.In the knitted version it appears more difficult to read, however that seems to be due to the number of vectors and limits on page size. All of the information is still included in the way it should be.

# Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r}
iso3uniq <- n_distinct(gapminder$iso3)
iso2uniq <- n_distinct(gapminder$iso2)
nameuniq <- n_distinct(gapminder$name)

gapminder %>% 
  summarize(iso3uniq=n_distinct(iso3), iso2uniq=n_distinct(iso2), nameuniq=n_distinct(name))
```


2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

_This doesn't seem accurate. I saw a note in the discord saying this question had a mistake_
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
_Note:_ Here I'm comparing just **_iso2_** and _name_
```{r, message=FALSE}
gapminder %>% 
  group_by(iso2, name) %>% 
  filter(!is.na(iso2)) %>% 
  summarize(n_distinct(name), n()) %>% 
  arrange(desc(n()))
```

This table shows the name and its correlated iso2-code, as well as the number of distinct names in that section (should be 1; appears to be 1 on each) and number of occurrences of that name. An example is Andorra. Its code is "AD", there is only one name included ("Andorra"), and it occurs 60 times within the data frame.

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entities are CHANISL and NLD CURACAO.
_Note:_ this one compares **_iso3_** and _name_
```{r, message=FALSE}
gapminder %>% 
  group_by(iso3, name) %>% 
  summarize(n_distinct(name), n()) %>% 
  arrange(desc(n()))
```
```{r}
gapminder %>% 
  filter(is.na(name)) %>% 
  group_by(name) %>% 
  filter(is.na(iso2)) %>% 
  distinct(iso3)
```


3. (2pt) What is the minimum and maximum year in these data?

_Note:_ in this data frame, year is shown in vector _time_

```{r}
#min year
gapminder %>% 
  arrange(time) %>% 
  select(time) %>% 
  head(1)
```
```{r}
#max year
gapminder %>% 
  filter(!is.na(time)) %>% 
  arrange(time) %>% 
  select(time) %>% 
  tail(1)
```

In the data, the minimum year is 1960, and the maximum year is 2019. This means that the data frame includes information of years 1960 through 2019.

# CO2 emissions (30pt)

Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

_Note:_ co2_PC is CO2 per capita

```{r}
gapminder %>% 
  group_by(time) %>% 
  summarize(sum(is.na(co2)), sum(is.na(co2_PC)))
```


2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.

_Note:_ I need to calculate top 3 co2 emitting countries (top 5 to add a few more)
```{r}
gapminder %>% 
  filter(!is.na(co2)) %>% 
  arrange(desc(co2)) %>% 
  head(5)
```
### This isn't it. fix it. idk how to plot evidently. i swear i'm trying


3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of then same countries. What does this figure suggest?


4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size. Hint: Americas 2016 should be 4.80.
```{r}
gapminder %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC)) %>% 
  summarize(mean(co2_PC))
```


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
```{r}
co2pc_1960 <- gapminder %>%
  filter(time==1960) %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC)) %>% 
  summarize(mean(co2_PC))
```
```{r}
co2pc_2016 <- gapminder %>%
  filter(time==2016) %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC)) %>% 
  summarize(mean(co2_PC))
```

fix```{r}
ggplot(co2pc_2016) +
  geom_col(x=region, y=co2pc_2016)
this```

### This isn't it. Fix it.

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).
```{r}
gapminder %>%
  select(time, region, name, co2_PC) %>%
  filter(time== 2016, !is.na(region)) %>%
  arrange(desc(co2_PC)) %>% 
  head(3)
```

```{r}
gapminder %>%
  select(time, region, name, co2_PC) %>%
  filter(time== 2016, !is.na(co2_PC)) %>%
  arrange(desc(co2_PC)) %>% 
  tail(3)
```

- Top 3 are Qatar, Unknown, and Trinidad and Tobego
- Bottom 3 are Burundi, Somalia, and Democratic Republic of the Congo

# GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC ).

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.


2. (4pt) Make a similar plot, but this time use 2019 data only.


3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?


4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.

Finally tell us how many hours did you spend on this PS.

